Sampling Strategies in Decisions from Experience

### 2022-05-04 manuscript.Rmd
Revised Structure

---
title             : |
  Sampling Strategies in Decisions from Experience

@@

abstract: |
  In decisions from experience, the probability to produce a higher outcome most of the time and the expected value are two latent properties that can be used to distinguish among the prospects of choice. This paper sketches a sampling and accumulation model that allows sampling strategies—i.e., the probability of sampling outcomes from different prospects in direct succession (switching probability)—to alter the degree to which evidence is accumulated for either one of these properties. A computational implementation of the model is used to simulate the respective sampling and accumulation process upfront choices between a safe and a two-outcome prospect, while varying switching probabilities and evidence thresholds. The simulation data shows that, depending on which latent property evidence is accumulated for and to which extent, respectively, choice patterns reflect various degrees of an as-if underweighting of rare outcomes or expected value maximization. Both choice patterns become more robust the more evidence is sampled and translate to characteristic parameter estimates for cumulative prospect theory’s value and weighting function.

@@

@@

What causes the description-experience gap?
A quite common notion is that of the mind being sensitive to the information provided by the small number of sampled outcomes that people tend to rely on in DFE [see @foxDecisionsExperienceSampling2006; @hertwigDecisionsExperienceEffect2004; @hertwigDecisionsExperienceWhy2010; @rakowBiasedSamplesNot2008; @wulffMetaanalyticReviewTwo2018].
According to this notion, small *samples*---i.e., sets of sampled outcomes---differ from descriptions in the information they actually provide about the prospects because they tend to underrepresent small-probability outcomes (Box 1).
This difference is then assumed to carry over to the decision-making process, eventually causing distinct choice patterns in DFD and DFE.
\newline

\noindent
\fbox{
    \parbox{\textwidth}{
        \textbf{Box 1.} 
        \textit{Reliance on Small Samples in DFE.}
        The binomial distribution of an outcome is the probability distribution of the number of times an outcome occurs in a sample.
        According to the \textit{de Moivre-Laplace theorem} (see, e.g., \protect\hyperlink{ref-georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015}{Georgii, 2015}), the binomial distribution can be approximated by a normal distribution given a large sample.
        That is, the relative relative frequency with which an outcome occurs in a large sample is predicted to correspond to the objective probability of the outcome most of the time and is not predicted to be lower more often than higher or vice versa.
        However, for small samples, the binomial distribution associated with a small-probability outcome is positive skewed, causing the relative frequency with which this outcome occurs in the sample to be smaller rather than larger than its objective probability.
        That is, small-probability outcomes tend to be underrepresented in small samples.
        An explanation for the as-if underweighting of rare outcomes pattern may thus build on the assumption that the mind is sensitive to the information provided by small samples (\protect\hyperlink{ref-hertwigDecisionsExperienceEffect2004}{Hertwig et al., 2004}; \protect\hyperlink{ref-plonskyRelianceSmallSamples2015}{Plonsky et al., 2015}; see also \protect\hyperlink{ref-erevChoicePredictionCompetition2010}{Erev et al., 2010}, \protect\hyperlink{ref-erevAnomaliesForecastsDescriptive2017}{2017}, for the performance of models assuming reliance on small samples in prediction competitions).
    }
}
\newline

Yet, the description-experience gap might not be solely caused by the reliance on a small number of sampled outcomes and the associated differences between the information underlying the decision-making process in DFD and DFE [see also @hauDecisionsExperienceStatistical2010; @hauDescriptionexperienceGapRisky2008; @ungemachAreProbabilitiesOverweighted2009; @wulffMetaanalyticReviewTwo2018, who show that the as-if underweighting of rare outcomes pattern may appear, even if the overall number of sampled outcomes is large and the relative frequencies closely resemble the latent objective probabilities].
Rather, the decision-making process itself may contribute to the gap.
That is, in DFE, people must ...

@hillsInformationSearchDecisions2010 argued that one contributing factor to the description-experience gap might be a potential link between what they considered two paradigmatic sampling and decision strategies (see Figure\ \@ref(fig:sampling-paradigm)).
That is, the authors assumed that in the sampling paradigm, people either switch prospects after each sampled outcome (piecewise sampling) to compare single outcomes over repeated rounds (round-wise decision), or they switch just once between prospects (comprehensive sampling) and make only one comparison between the means across all sampled outcomes (summary decision).
In fact, @hillsInformationSearchDecisions2010 provided tentative evidence for such a coupling of sampling and decision strategies by finding that the choices of people who switched frequently between prospects during the sampling phase were better predicted by a round-wise strategy (rather than a summary strategy), and that this predictive pattern reversed for people who switched infrequently.^[<!--Open footnote-->
Note that in their analyses of empirical sampling data,  @hillsInformationSearchDecisions2010 used the number of switches between prospects relative to the overall number of sampled outcomes from both prospects (switching frequency) as an approximate measure for the two proposed sampling strategies, and acknowledged that "many [sampling] strategies will fall on the continuum in between [the two paradigmatic strategies]" (p. 1788).
<!--Close footnote-->]
Using the round-wise strategy may translate to an as-if underweighting of rare events pattern and the description-experience gap because the strategy selects the prospect that wins a majority of comparisons, while small-probability outcomes are predicted not to be considered in most of these comparisons.

@@

<!-- Problems of Hills and Hertwig -->

@@

<!--To model the process by which people arrive at the final choice, one must explicate a number of assumptions, e.g.: How do people decide when to switch between prospects during the sampling phase; how do they integrate and process the information obtained from the sampled outcomes; and, because they can sample as much as they want, how do they decide to stop sampling.
To approach these questions, the model presented in this paper integrates the proposed link between sampling and decision strategies into the computational framework of sequential sampling models.-->
Now, the current paper presents a sampling and accumulation model that provides an algorithmic link between sampling and decision strategies, and uses the model to discuss how the mind may rely on sampling strategies to alter which differences in the latent properties of prospects evidence is accumulated for.
More specifically, the proposed model assumes an accumulation process over a sequence of mean comparisons, with switching probabilities---i.e., the probability of sampling
outcomes from different prospects in direct succession---determining the size of the samples underlying each mean comparison and thresholds determining how many comparisons a prospect must win in order to be chosen.
The model is such that when switching probabilities are high, the accumulation process unfolds over mean comparisons based on small samples, thus resembling the round-wise strategy and accumulating evidence for differences in the probability to produce a higher outcome most of the time.
Notably, in this case, small-probability outcomes are predicted to contribute to the majority of mean comparisons less than would be objectively warranted, which can translate to an as-if underweighting of rare outcomes, irrespective of what the thresholds require the number of won comparisons to be.
<!--Accordingly, while increases in thresholds---and thus in the number of sampled outcomes distributed over multiple small samples, each of which is considered---allow to choose with increasing accuracy the prospect that is more likely to return a higher outcome, the resulting choices can produce an as-if underweighting of rare outcomes.-->
In turn, for low switching probabilities, the accumulation process unfolds over mean comparisons based on large samples, thus resembling the summary strategy---for low thresholds, respectively---and accumulating evidence for differences in the EV.

To substantiate the discussion, the model is used to simulate the sampling and accumulation process for choices between a safe and a two-outcome prospect, while systematically combining different switching probabilities and thresholds.
Considering choice rates and parameter estimates of cumulative prospect theory [@tverskyAdvancesProspectTheory1992], the computational analysis shows that the choices converge to the solutions of EV maximization for low switching probabilities, but take the form of an as-if underweighting of rare outcomes for high switching probabilities.
Importantly, the analysis also shows that these distinct choice patterns become more robust with increasing thresholds.
Before presenting the model and the computational analysis, the following section provides a definition of prospects.

# Definition of Prospects

# Sampling Strategies and the Accumulation of Evidence

@@

## Model

We consider the sampling paradigm and a choice between two prospects denoting the distributions $P'^X$ and $P'^Y$ of the discrete random variables $X$ and $Y$, respectively.
The model assumes that the sampling process starts at random with a stochastic process on one of the two random variables.
Specifically, it is assumed that an agent starts with equal probability to sample from one of the prospects, say $P'^X$, and generates a sequence of random variables $X_1, X_2,...$, where the mean over the sequence is computed or updated with each new outcome that is sampled.
The respective stochastic process $\{S^X_n\}_{n \in \mathbb{N}}$ is defined by
$$
S^X_n = \frac{1}{n} \sum_{t = 1}^n X_t, \quad n \in \mathbb{N}
\; ,
\tag{7}
$$
and terminates as soon as an outcome is sampled from the other prospect.
The switching probability $\psi \in (0,1]$, which is assumed to be adopted prior the start of the information-processing chain and is fixed throughout, controls the probability with which outcomes from different prospects are sampled in direct succession.
In other words, the switching probability is the probability with which the stochastic process on the current random variable stops and a new stochastic process on the other random variable starts with the subsequently sampled outcome.
To implement the accumulation process, the model assumes that each time after a new stochastic process was started and terminated on both random variables, their respective values are compared, with the prospect underlying the stochastic process with the greater final value receiving a round win.
Accordingly, for the prospect $P'^X$, this stage is modeled as the mapping
$$
\begin{aligned}
  Z^X : \mathbb{R} &\to \mathbb{N} \\
  S^X_n - S^Y_n &\mapsto 
  \begin{cases}
     1 &\text{if} \quad  S^X_n > S^Y_n, \\
     0 &\text{else}
     \; ,
  \end{cases}
\end{aligned}
\tag{8}
$$
where $Z^X$ is a random variable that takes only the two values $0$ and $1$, denoting a lost or won mean comparison, respectively. 
The model assumes that a round consisting of a pair of stochastic processes---i.e., one for each prospect---and the subsequent comparison of their final values is repeated until the sum of round wins for one of the two prospects reaches a threshold $\theta \in \mathbb{N}$.
Similar to the switching probability, it is assumed that the threshold is adopted prior the start of the information-processing chain and is fixed throughout.
For $P'^X$, then, we obtain a sequence of independent comparisons $Z^X_1, Z^X_2, ...$ and model the respective evidence accumulation process as a random walk $\{D^X_m\}_{m \in \mathbb{N}}$ defined by 
$$
D^X_m = \sum_{i = 1}^m Z^X_i, \quad m \in \mathbb{N}
\;,
\tag{9}
$$
resembling a dynamic decision variable.
Note that since (8) describes a Bernoulli trial, the random walk in (9) is based on a Bernoulli process.

@@

>>>>

---
title             : |
  Sampling Strategies in Decisions from Experience: Property Representation and Probability Weighting Approximation

@@

abstract: |
  Add abstract

@@

<!-- Introduction -->

@@

What causes the description-experience gap?
A quite common notion is that of the mind being sensitive to the information provided by the small number of sampled outcomes that people tend to rely on in DFE [see @foxDecisionsExperienceSampling2006; @hertwigDecisionsExperienceEffect2004; @hertwigDecisionsExperienceWhy2010; @rakowBiasedSamplesNot2008; @wulffMetaanalyticReviewTwo2018].
Specifically, small *samples*---i.e., sets of sampled outcomes---often differ from descriptions in the information they actually provide about the prospects because they tend to underrepresent small-probability outcomes (Box \hyperlink{box}{1}, see also for illustrative purposes the `code example`, where the objective probability of outcome `5` is `.2`, and the sampled relative frequency is `.1`).
\newline

\noindent
\fbox{
    \parbox{\textwidth}{
        \textbf{\hypertarget{box}{Box 1.}} 
        \textit{Reliance on Small Samples in DFE.}
        The binomial distribution of an outcome is the probability distribution of the number of times an outcome occurs in a sample.
        According to the \textit{de Moivre-Laplace theorem} (see, e.g., \protect\hyperlink{ref-georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015}{Georgii, 2015}), the binomial distribution can be approximated by a symmetric normal distribution given a large sample.
        That is, the relative frequency with which an outcome occurs in a large sample is predicted to correspond to the objective probability of the outcome most of the time and is not predicted to be lower more often than higher or vice versa.
        However, for small samples, the binomial distribution associated with a small-probability outcome is positive skewed, causing the relative frequency with which this outcome occurs in the sample to be smaller rather than larger than its objective probability.
        That is, small-probability outcomes tend to be underrepresented in small samples.
        An explanation for the as-if underweighting of rare outcomes pattern may thus build on the assumption that the mind is sensitive to the information provided by small samples (\protect\hyperlink{ref-hertwigDecisionsExperienceEffect2004}{Hertwig et al., 2004}; \protect\hyperlink{ref-plonskyRelianceSmallSamples2015}{Plonsky et al., 2015}; see also \protect\hyperlink{ref-erevChoicePredictionCompetition2010}{Erev et al., 2010}, \protect\hyperlink{ref-erevAnomaliesForecastsDescriptive2017}{2017}, for the performance of models assuming reliance on small samples in prediction competitions).
    }
}
\newline

\noindent
This distortion of the sampled information (sampling error) is then assumed to carry over to the decision-making process, eventually causing distinct choice patterns in DFD and DFE.

Yet, the description-experience gap might not be solely caused by the reliance on a small number of sampled outcomes and the associated differences in the information underlying the decision-making processes in DFD and DFE [see also @hauDecisionsExperienceStatistical2010; @hauDescriptionexperienceGapRisky2008; @ungemachAreProbabilitiesOverweighted2009; @wulffMetaanalyticReviewTwo2018, who show that the as-if underweighting of rare outcomes pattern may appear, even if the overall number of sampled outcomes is large and the relative frequencies closely resemble the latent objective probabilities].
Rather, the specifics of the decision-making processes themselves may contribute to the gap.
To this end, the current paper models how the mind might exercise the capacity of making decisions from experience in the sampling paradigm and examines the choice patterns that result from the modeled decision-making processes.
More specifically, controlling for sampling error---i.e., considering the sampled relative frequencies rather than the latent objective probabilities---, this paper investigates in a computational analysis whether and how the interplay of previously proposed sampling and decision strategies [see @hillsInformationSearchDecisions2010, see also below] shapes DFE, and comes to imply choice patterns and functional forms in cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992] that are consistent with an as-if underweighting of rare outcomes pattern.^[<!--Open footnote-->
Note that CPT's value function is also considered. However, the focus of this paper is on the weighting function, as it is commonly used to describe an as-if overweighting and underweighting of rare outcomes, as well as the description-experience gap [see, e.g., @kellenHowVariantAre2016; @regenwetterConstructbehaviorGapBehavioral2017; but see @hertwigConstructbehaviorGapDescription2018].
<!--Close footnote-->]

The remainder of the paper is structured as follows:
[...]

<!-- To substantiate the discussion, the model is used to simulate the sampling and accumulation process for choices between a safe and a two-outcome prospect, while systematically combining different switching probabilities and thresholds.
Considering choice rates and parameter estimates of cumulative prospect theory [@tverskyAdvancesProspectTheory1992], the computational analysis shows that the choices converge to the solutions of EV maximization for low switching probabilities, but take the form of an as-if underweighting of rare outcomes for high switching probabilities.
Importantly, the analysis also shows that these distinct choice patterns become more robust with increasing thresholds.
Before presenting the model and the computational analysis, the following section provides a definition of prospects.-->

# Explaining Decision Making in the Sampling Paradigm

<!-- Formal Definition of Prospects / Notation -->

<!-- Modeling steps and assumptions -->

To model the process by which people arrive at the final choice in the sampling paradigm, one must explicate a number of assumptions, e.g.:

* How does the mind decide when to switch between prospects during the sampling phase?

* How does the mind decide when to stop sampling?

* How does the mind integrate and evaluate the information obtained from the sampled outcomes?

To approach the above questions, this paper presents a model that integrates two proposed sampling and decision strategies [see @hillsInformationSearchDecisions2010] into the computational framework of sequential sampling models of decision-making [see, e.g., @markantModelingChoiceSearch2015].

# Sampling and Decision Strategies

Specifically, @hillsInformationSearchDecisions2010 argued that one contributing factor to the description-experience gap might be a potential link between what they considered two paradigmatic sampling and decision strategies (see Figure\ \@ref(fig:sampling-paradigm)).
That is, the authors assumed that in the sampling paradigm, people either switch prospects after each sampled outcome (piecewise sampling) to compare single outcomes over repeated rounds (round-wise decision), or they switch just once between prospects (comprehensive sampling) and make only one comparison between the means across all sampled outcomes (summary decision).
In fact, @hillsInformationSearchDecisions2010 provided tentative evidence for such a coupling of sampling and decision strategies by finding that the choices of people who switched frequently between prospects during the sampling phase were better predicted by a round-wise strategy (rather than a summary strategy), and that this predictive pattern reversed for people who switched infrequently.^[<!--Open footnote-->
Note that in their analyses of empirical sampling data,  @hillsInformationSearchDecisions2010 used the number of switches between prospects relative to the overall number of sampled outcomes from both prospects (switching frequency) as an approximate measure for the two proposed sampling strategies, and acknowledged that "many [sampling] strategies will fall on the continuum in between [the two paradigmatic strategies]" (p. 1788).
<!--Close footnote-->]
Using the round-wise strategy may translate to an as-if underweighting of rare events pattern and the description-experience gap because the strategy selects the prospect that wins a majority of comparisons, while small-probability outcomes are predicted not to be considered in most of these comparisons.

@@

@@

<!-- Critique -->

# Current Approach: Sampling Strategies and the Accumulation of Different Types of Evidence

<!-- How to overcome the problems + Evidence accumulation -->

Now, the current paper presents a sampling and accumulation model that provides an algorithmic link between sampling and decision strategies, and uses the model to discuss how the mind may rely on sampling strategies to alter which differences in the latent properties of prospects evidence is accumulated for.
More specifically, the proposed model assumes an accumulation process over a sequence of mean comparisons, with switching probabilities---i.e., the probability of sampling
outcomes from different prospects in direct succession---determining the size of the samples underlying each mean comparison and thresholds determining how many comparisons a prospect must win in order to be chosen.
The model is such that when switching probabilities are high, the accumulation process unfolds over mean comparisons based on small samples, thus resembling the round-wise strategy and accumulating evidence for differences in the probability to produce a higher outcome most of the time.
Notably, in this case, small-probability outcomes are predicted to contribute to the majority of mean comparisons less than would be objectively warranted, which can translate to an as-if underweighting of rare outcomes, irrespective of what the thresholds require the number of won comparisons to be.
<!--Accordingly, while increases in thresholds---and thus in the number of sampled outcomes distributed over multiple small samples, each of which is considered---allow to choose with increasing accuracy the prospect that is more likely to return a higher outcome, the resulting choices can produce an as-if underweighting of rare outcomes.-->
In turn, for low switching probabilities, the accumulation process unfolds over mean comparisons based on large samples, thus resembling the summary strategy---for low thresholds, respectively---and accumulating evidence for differences in the EV.

@@

## Models

We consider the sampling paradigm and a choice between two prospects denoting the distributions $P'^X$ and $P'^Y$ of the discrete random variables $X$ and $Y$, respectively.

### Round-Wise Integration Model

The model assumes that the sampling process starts at random with a stochastic process on one of the two random variables.
Specifically, it is assumed that an agent starts with equal probability to sample from one of the prospects, say $P'^X$, and generates a sequence of random variables $X_1, X_2,...$, where the mean over the sequence is computed or updated with each new outcome that is sampled.
The respective stochastic process $\{S^X_n\}_{n \in \mathbb{N}}$ is defined by
$$
S^X_n = \frac{1}{n} \sum_{t = 1}^n X_t, \quad n \in \mathbb{N}
\; ,
\tag{7}
$$
and terminates as soon as an outcome is sampled from the other prospect.
The switching probability $\psi \in (0,1]$, which is assumed to be adopted prior the start of the information-processing chain and is fixed throughout, controls the probability with which outcomes from different prospects are sampled in direct succession.
In other words, the switching probability is the probability with which the stochastic process on the current random variable stops and a new stochastic process on the other random variable starts with the subsequently sampled outcome.
To implement the accumulation process, the model assumes that each time after a new stochastic process was started and terminated on both random variables, their respective values are compared, with the prospect underlying the stochastic process with the greater final value receiving a round win.
Accordingly, for the prospect $P'^X$, this stage is modeled as the mapping
$$
\begin{aligned}
  Z^X : \mathbb{R} &\to \mathbb{N} \\
  S^X_n - S^Y_n &\mapsto 
  \begin{cases}
     1 &\text{if} \quad  S^X_n > S^Y_n, \\
     0 &\text{else}
     \; ,
  \end{cases}
\end{aligned}
\tag{8}
$$
where $Z^X$ is a random variable that takes only the two values $0$ and $1$, denoting a lost or won mean comparison, respectively. 
The model assumes that a round consisting of a pair of stochastic processes---i.e., one for each prospect---and the subsequent comparison of their final values is repeated until the sum of round wins for one of the two prospects reaches a threshold $\theta \in \mathbb{N}$.
Similar to the switching probability, it is assumed that the threshold is adopted prior the start of the information-processing chain and is fixed throughout.
For $P'^X$, then, we obtain a sequence of independent comparisons $Z^X_1, Z^X_2, ...$ and model the respective evidence accumulation process as a random walk $\{D^X_m\}_{m \in \mathbb{N}}$ defined by 
$$
D^X_m = \sum_{i = 1}^m Z^X_i, \quad m \in \mathbb{N}
\;,
\tag{9}
$$
resembling a dynamic decision variable.
Note that since (8) describes a Bernoulli trial, the random walk in (9) is based on a Bernoulli process.

@@

### Summary Integration Model
